---
title: "EVPI for external validation"
author: "Sadatsafavi et al"
date: "2022.08.19"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source('helper_functions.R')
```

# Data preparation and processing

```{r,message=F}
library(voipred) # https://github.com/resplab/voipred
library(predtools) 
library(rms)
library(mvtnorm)
library(pROC)
library(sqldf)
set.seed(123) 

settings <- list(
  dev_sample_size = Inf,
  val_sample_size = 500,
  n_sim=10000,
  Bayesian_bootstrap =T,
  zs=c(0.01,0.05,0.1,0.02),
  max_x_evpi_plot = 0.49   #For plots of evpi 
)

data(gusto)
gusto$kill <- (as.numeric(gusto$Killip)>1)*1
gusto$Y <- gusto$day30
data_us <- gusto[gusto$regl %in% c(1, 7, 9, 10, 11, 12, 14, 15),]
data_other <- gusto[!gusto$regl %in% c(1, 7, 9, 10, 11, 12, 14, 15),]

if(is.infinite(settings$dev_sample_size))
{
  dev_data <- data_other
}else
{
  dev_data <- data_other[sample(1:(dim(data_other)[1]),settings$dev_sample_size,F),]
}

if(is.infinite(settings$val_sample_size))
{
  val_data <- data_us
}else
{
  #data is for external validation
  val_data <- data_us[sample(1:(dim(data_us)[1]),settings$val_sample_size,F),]  
}

model <- glm(Y ~ age + miloc + pmi + kill + pmin(sysbp,100) + pulse, data=dev_data, family=binomial(link="logit"))

pi <- predict(model, type="response", newdata=val_data)
pi <- 1/(1+exp(0-(log(pi/(1-pi)))))
val_data$pi <- pi

zs <- (c(as.vector(quantile(pi,(1:99)/100)),settings$zs))
main_index <- length(zs)
```

# Candidate model's coefficients and c-statistic
```{R}
coefficients(model)
pROC::auc(val_data$Y,val_data$pi)
```

# Figure 1: Decision curve ($\hat{NB}$) of the candidate model (blue), treating all (black oblique line), and treating none (black horizontal line) in the validation sample A bootstrapped 95% confidence interval is given for the candidate model and treating all strategies (gray curves).

```{r}
n <- dim(val_data)[1]
NBh_model <- NBh_all <- rep(0,length(zs))
for(i in 1:length(zs))
{
  NBh_model[i] <- mean((val_data$pi>zs[i])*(val_data$Y - (1-val_data$Y)*zs[i]/(1-zs[i])))
  NBh_all[i] <- mean((val_data$Y - (1-val_data$Y)*zs[i]/(1-zs[i])))
}

# Uncertainty around it
bs_NBh_model <- bs_NBh_all <- matrix(0, nrow=settings$n_sim, ncol=length(zs))

for(i_sim in 1:settings$n_sim)
{
  w <- voipred:::bootstrap(n,settings$Bayesian_bootstrap)
  
  for(i in 1:length(zs))
  {
    bs_NBh_model[i_sim,i] <- sum(w*(val_data$pi>zs[i])*(val_data$Y - (1-val_data$Y)*zs[i]/(1-zs[i])))/n
    bs_NBh_all[i_sim,i] <- sum(w*(val_data$Y - (1-val_data$Y)*zs[i]/(1-zs[i])))/n
  }
}

NBh_all <- colMeans(bs_NBh_all)
NBh_model <- colMeans(bs_NBh_model)

max_y <- max(0,NBh_all[1:99],NBh_model[1:99])
plot(zs[1:99], NBh_all[1:99], col='black', type='l', 
     ylim=c(0,max_y), xlab="Risk threshold", ylab="Net benefit")
lines(zs[1:99], zs[1:99]*0, col='black', type='l')
lines(zs[1:99], NBh_model[1:99], col='blue', type='l')
ci_model <- apply(bs_NBh_model[,1:99],MARGIN = 2,FUN = quantile, c(0.025,0.975))
lines(zs[1:99],ci_model[1,], type='l', col='gray')
lines(zs[1:99],ci_model[2,], type='l', col='gray')
```



# Figure 2: Histogram of the incremental NB of the model ($\hat{NB}_{model}$-maxâ¡{0,$\hat{NB}_{all}$}) based on 10,000 bootstraps.

```{r}
hist(bs_NBh_model[,main_index]-bs_NBh_all[,main_index], xlab="Incremental NB", ylab="Density", main="")
```

# Values of decision at threshold values of interest:
```{r}
rbind(z=settings$zs,NB_model=NBh_model[-(1:99)],NB_all=NBh_all[-(1:99)],INB=NBh_model[-(1:99)]-NBh_all[-(1:99)])
```


# Figure 3: Validation EVPI for the case study as a function of thresholds for the Bayesian bootstrap (red), ordinary bootstrap (blue), and asymptotic method (orange)
```{r}
zs_evpi <- seq(0,settings$max_x_evpi_plot,by=0.01)
res_bb <- evpi_val(Y = val_data$Y,
                   pi = val_data$pi,
                   method='bayesian_bootstrap',
                   zs = zs_evpi,
                   n_sim=settings$n_sim)

res_ob <- evpi_val(Y = val_data$Y,
                   pi = val_data$pi,
                   method='bootstrap',
                   zs = zs_evpi,
                   n_sim=settings$n_sim)  

res_as <- evpi_val(Y = val_data$Y,
                   pi = val_data$pi,
                   method='asymptotic',
                   zs = zs_evpi,
                   n_sim=settings$n_sim)

max_y <- max(c(res_bb$EVPIv,res_ob$EVPIv,res_as$EVPIv))
plot(zs_evpi,res_bb$EVPIv,type='l', col='red', lwd=2, 
     ylim=c(0,max_y), xlim=c(0,0.20), xlab="Risk threshold", ylab="EVPI")
lines(zs_evpi,res_ob$EVPIv,type='l', col='blue', lwd=2)
lines(zs_evpi,res_as$EVPIv,type='l', col='orange', lwd=2)
```

# Figure 5: EVPI values across the range of sample sizes for the three computation methods and threshold values of 0.01 (top left), 0.02 (top right), 0.05 (bottom left), and 0.10 (bottom-right). Red: Bayesian bootstrap; blue: ordinary bootstrap; orange: asymptotic method

```{r}
sim_by_size <- function(n_sim=1000, sample_sizes=c(250, 500, 1000,2000,4000,8000,16000,Inf),zs=c(0.01,0.02,0.05,0.1))
  {
  set.seed(1)
  out <- data.frame(method=character(), sample_size=integer())
  for(i in 1:length(zs))
  {
    out[paste0('val',i)] <- double()
  }
  
  index <- 1
  
  for(s in 1:length(sample_sizes))
  {
    sample_size <- sample_sizes[s]
    cat(sample_size)
    res_bb <- res_ob <- res_ll <- rep(0,length(zs))
    
    for(i in 1:n_sim)
    {
      #set.seed(i)
      # cat('.')
      if(is.infinite(sample_size))
      {
        this_data <- data_us
        this_data$pi <- predict(model,newdata = this_data, type='response')
        sample_size <- dim(this_data)[1]
      }
      else
      {
        repeat
        {
          this_data <- data_us[sample(1:(dim(data_us)[1]),sample_size,F),]  #data is for external validation
          this_data$pi <- predict(model,newdata = this_data, type='response')
          if(min(this_data$pi)<min(zs))
          {
            break;
          }
          else
          {
            cat('bad')
          }
        }
      }
      
      bad <- F
      
      tmp <- evpi_val(Y = this_data$Y,pi=this_data$pi,
                      method="bayesian_bootstrap", zs = zs)      
      if(is.null(tmp)) bad <- T
      out[index,'method'] <- "BB"; out[index,'sample_size']<-sample_size; out[index,c('val1','val2','val3','val4')] <- tmp$EVPIv
      index <- index+1
      
      tmp <- evpi_val(Y = this_data$Y,pi=this_data$pi,
                      method="bootstrap", zs = zs)  
      if(is.null(tmp)) bad <- T
      out[index,'method'] <- "OB"; out[index,'sample_size']<-sample_size; out[index,c('val1','val2','val3','val4')] <- tmp$EVPIv
      index <- index+1
      
      tmp <- evpi_val(Y = this_data$Y,pi=this_data$pi,
                method="asymptotic", zs = zs)  
      if(is.null(tmp)) bad <- T
      out[index,'method'] <- "asy"; out[index,'sample_size']<-sample_size; out[index,c('val1','val2','val3','val4')] <- tmp$EVPIv
      index <- index+1
      
      if(bad)
      {
        index <- index -3
        i <- i-1
        message("bad")
      }
    }
  }
  
  res <<- out
  
  out
}
res <- sim_by_size(n_sim=1000)
pres <- sqldf("SELECT method, sample_size, AVG(val1) AS val1, AVG(val2) AS val2 , AVG(val3) AS val3 , AVG(val4) AS val4 from res GROUP BY method, sample_size")

max_y <- max(pres[which(pres$method=="BB"),]$val1,pres[which(pres$method=="OB"),]$val1,pres[which(pres$method=="asy"),]$val1)
plot(pres[which(pres$method=="BB"),]$sample_size,pres[which(pres$method=="BB"),]$val1,type='l', xlab = "Sample size", ylab="EVPI", col='red', lwd=2, ylim=c(0,max_y),main="z=0.01")
lines(pres[which(pres$method=="OB"),]$sample_size,pres[which(pres$method=="OB"),]$val1,col='blue', lwd=2)
lines(pres[which(pres$method=="asy"),]$sample_size,pres[which(pres$method=="asy"),]$val1,col='orange', lwd=2)

max_y <- max(pres[which(pres$method=="BB"),]$val2,pres[which(pres$method=="OB"),]$val2,pres[which(pres$method=="asy"),]$val2)
plot(pres[which(pres$method=="BB"),]$sample_size,pres[which(pres$method=="BB"),]$val2,type='l', xlab = "Sample size", ylab="EVPI", col='red', lwd=2, xlim=c(0,9000), ylim=c(0,max_y),main="z=0.02")
lines(pres[which(pres$method=="OB"),]$sample_size,pres[which(pres$method=="OB"),]$val2,col='blue', lwd=2)
lines(pres[which(pres$method=="asy"),]$sample_size,pres[which(pres$method=="asy"),]$val2,col='orange', lwd=2)

max_y <- max(pres[which(pres$method=="BB"),]$val3,pres[which(pres$method=="OB"),]$val3,pres[which(pres$method=="asy"),]$val3)
plot(pres[which(pres$method=="BB"),]$sample_size,pres[which(pres$method=="BB"),]$val3,type='l', xlab = "Sample size", ylab="EVPI", col='red', lwd=2, xlim=c(0,3000), ylim=c(0,max_y),main="z=0.05")
lines(pres[which(pres$method=="OB"),]$sample_size,pres[which(pres$method=="OB"),]$val3,col='blue', lwd=2)
lines(pres[which(pres$method=="asy"),]$sample_size,pres[which(pres$method=="asy"),]$val3,col='orange', lwd=2)

max_y <- max(pres[which(pres$method=="BB"),]$val4,pres[which(pres$method=="OB"),]$val4,pres[which(pres$method=="asy"),]$val4)
plot(pres[which(pres$method=="BB"),]$sample_size,pres[which(pres$method=="BB"),]$val4,type='l', xlab = "Sample size", ylab="EVPI", col='red', lwd=2, xlim=c(0,3000), ylim=c(0,max_y),main="z=0.10")
lines(pres[which(pres$method=="OB"),]$sample_size,pres[which(pres$method=="OB"),]$val4,col='blue', lwd=2)
lines(pres[which(pres$method=="asy"),]$sample_size,pres[which(pres$method=="asy"),]$val4,col='orange', lwd=2)
```

# P(useful)
```{r}
zs_index <- match(settings$zs,zs_evpi)
max_y <- max(c(res_bb$p_useful,res_ob$p_useful,res_as$p_useful))
plot(zs_evpi,res_bb$p_useful,type='l', col='red', lwd=2, ylim=c(0,max_y), xlab="Risk threshold", ylab="P(useful)")
lines(zs_evpi,res_ob$p_useful,type='l', col='blue', lwd=2)
rbind(z=settings$zs,
      p_useful_bb=res_bb$p_useful[zs_index],
      p_useful_ob=res_ob$p_useful[zs_index])
```


# EVPI at specified threhsolds
```{r}
rbind(z=settings$zs,
      EVPIv_bb=res_bb$EVPIv[zs_index],
      EVPIv_ob=res_ob$EVPIv[zs_index],
      EVPIv_as=res_as$EVPIv[zs_index])
```